Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-02-09 alloscore_manuscript/alloscore-application-shorter.Rnw
intro rewrite

Infectious disease forecasting models have emerged as important tools in public health outbreak response. Predictions of disease dynamics have been used to inform decision making about a wide variety of measures to reduce disease spread and/or mitigate the severity of disease outcomes. For example, estimates of the onset time of the flu season have been used to aid national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola and diptheria dynamics have been used to allocate resources \citep{meltzer2014estimating, rainisch2015regional,camacho2015-ebola-bed,finger_real-time_2019}. \cite{bertsimas2021predictionsCOVID} developed tools to guide decision making from infectious disease forecasts, which have been used to inform allocation of limited resources such as ventilators, ICU capacity, and vaccine supply. Models developed by \cite{fox_real-time_2022} have been used to inform resource and care site planning, as well as community guidelines for masking, traveling, dining and shopping (University of Texas, 2022)\nocite{utnews2022}.
%In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response.

In the wake of the COVID-19 pandemic, researchers have been calling for evaluating infectious disease forecasts through their impact on policy \citep{marshall2023predictions, bilinski_adaptive_2023}.
Historically, the quality of disease forecasts has typically been measured using scores such as root mean square error (RMSE) \citep{papastefanopoulos2020covid}, log score \citep{mcgowan_collaborative_2019}, or weighted interval score (WIS) \citep{bracher2021evaluating}.  These scores can be interpreted as measuring how well the forecast corresponds to the observed disease outcome.
However, as we show in this paper, these scores often do not clearly measure the value of a forecast in the context of a specific decision.
While these ideas have received little attention in the context of infectious disease forecasting, the results are well known from other fields, such as economics and finance, supply chain management, and meteorology.

In decision making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to (see details in Section \ref{sec:methods.detailed.decisiontheory}).
%There is a large history of literature applying these ideas to obtain measures of the value of forecasts that are tied to a decision making context, primarily in fields such as economics and finance, supply chain management, and meteorology.
We refer the reader to \cite{yardley2021utility_cost_forecasts} for a general overview, and to \cite{pesaran2002decision_based_eval} and \cite{murphy1993whatisagoodforecast} for discussions focused on applications to economics and meteorology, respectively.
%In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact \cite[e.g.,][]{catt2007assessingcostofforecasterror, petropoulos2019inventoryperformanceforecasting, palmer2002economic, pappenberger2015monetarybenefitfloodwarnings}.
In this framework, a forecast has value if it leads to decisions with low total costs.
As one example,
% in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while
in meteorology, policy makers need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages \citep{palmer2002economic, pappenberger2015monetarybenefitfloodwarnings}.
Across many fields, analyses have consistently found that common measures of forecast accuracy do not necessarily correspond directly to measures of the value of forecasts as an input to decision making \cite[e.g.,][]{leitch1991economicForecastEval, murphy1993whatisagoodforecast, cenesizoglu2012returnPredictionEconValue}.

We are aware of only a limited body of work that explicitly attempts to measure the value of infectious disease forecasts through their impact on policy. For example, \cite{ioannidis2022forecastingCOVIDfailed} discuss the possible negative consequences of inaccurate forecasts of infectious disease, but do not attempt to quantify the utility or loss incurred as a result of those forecasts. \cite{bilinski_adaptive_2023} explore ways in which policymaker preferences could inform risk thresholds for predictive models using a framework for measuring the costs and losses associated with taking an action that is similar to methods that have been used in meteorology and elsewhere. \cite{marshall2023predictions} develop a forecast scoring rule that is informally motivated by utility considerations, but the score is not derived from a decision-theoretic set up. Separately, there is a thread of literature that quantifies the link between infectious disease modeling and policy making outside of a forecasting context. As an example, \cite{Probert2016decisionMakingFootMouth} develop measures of the cost of actions designed to control a hypothetical outbreak of foot-and-mouth disease and use this framework to explore policy recommendations from a variety of simulation-based projection models.

In practice, probabilistic infectious disease forecasts have most often been made for observations that emerge from public health surveillance systems and have typically been evaluated with common scoring rules that have implementations available in existing software packages.
For example, seasonal influenza forecasts in the US and dengue forecasts for Peru and Puerto Rico targeted public health surveillance measures of incidence over time and space, and used log score and mean absolute error to evaluate forecast skill \citep{mcgowan_collaborative_2019,reich_collaborative_2019,johansson_open_2019}.
Pandemic COVID-19 forecasts of observed cases, hospitalizations and/or deaths in the US and Europe, as reported by municipal, state, or federal surveillance systems, were evaluated using the weighted interval score (WIS, which is an approximation of the continuous ranked probability score, or CRPS), and prediction interval coverage \citep{cramer_evaluation_2022,fox_real-time_2022,sherratt2023predictive}.
Similarly, CRPS was also used to assess probabilistic forecasts of dengue incidence at the district level in Vietnam \citep{colon-gonzalez_probabilistic_2021}.
While some of these scores can be interpreted through the lens of decision theory, and all of the application-specific papers cited above have benefited from direct collaboration with public health agencies, none of them make explicit connections between forecast evaluation and how a forecast was used in practice.

In this work, we begin to fill this gap between the ways that infectious disease forecasts have traditionally been evaluated and the ways that they have been used to support public health policy.
We consider a setting in which forecasts are used to help determine the allocation of a limited quantity of medical supplies across multiple regions.
In section~\ref{sec:methods} of the paper, we define a new forecast scoring rule --- the {\em allocation score} --- that evaluates forecasts based on how beneficial resource allocations derived from them would turn out to be.
%Unlike the traditional measures of forecast skill such as WIS, CRPS, and log score, the proposed allocation score measures the suitability of a forecast as an input to this specific decision making task.
In section \ref{sec:application}, we present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US leading up to the Omicron wave in winter 2022.
This analysis is ``synthetic'' in that it does not correspond to an actual analysis that supported decision making in real time.
However, the framework described in this paper corresponds to real-world decisions that must be made by public health administrators around the globe, and could be adapted in the future for real-time situations such as informing bed demand, ventilator usage, or doses needed for specific medications.
%For example, forecasts for districts in Sierra Leone of bed demand to care for patients with Ebola was the subject of a real-time modeling study in late 2014 and early 2015 \citep{camacho2015-ebola-bed}.
%And, in 2020, a model developed by an academic research group turned predictions of COVID-19 hospitalizations into estimates of ventilator usage and shortages. This framework was used by the Hartford HealthCare system in Connecticut ``to align ventilator supply with projected demand at a time where the [COVID-19] pandemic was on the rise'' \citep{bertsimas2021predictionsCOVID}.
%These examples illustrate the potential for forecasts to inform decisions about how to allocate limited supplies such as temporary hospital beds, ventilators, personal protective equipment, or other supplies that are known to be effective at reducing transmission or severity of disease.
%However, we emphasize again that these studies did not take the step of evaluating forecasts based on the quality of the allocation decisions that they supported or could have been used to support.

>>>>

Infectious disease forecasting models have emerged as important tools in public health outbreak response. The predictions they provide increasingly inform decisions regarding a wide variety of countermeasures intended to reduce transmission and mitigate the severity of disease outcomes. For example, estimates of the onset time of the flu season have been used in developing national vaccination strategies \citep{igboh2023timing}, and forecasts of Ebola and diptheria dynamics have been made with the clearly stated goal of helping local 
public health workers choose the timing and location of interventions in settings where resources are severely constrained \citep{meltzer2014estimating, rainisch2015regional,camacho2015-ebola-bed,finger_real-time_2019}. More recently, in the context of the outbreak of COVID-19 across the US, \cite{bertsimas2021predictionsCOVID} explicitly based triage decision tools for the interstate reallocation
of ventilators and ICU capacity on forecast inputs and used these forecasts as well to recommend vaccine trial sites to a 
major trial sponsor. \cite{fox_real-time_2022} similarly used predictive models to inform intrastate resource and care site planning, as well as local community guidelines for masking, traveling, dining and shopping (University of Texas, 2022)\nocite{utnews2022}.
%In April of 2022, the Centers for Disease Control and Prevention (CDC) announced the launch of the Center for Forecasting and Outbreak Analytics (CFA) to translate disease forecasts into decision making (CDC, 2022)\nocite{cdc2022cfa}, indicating that this has been identified as an important direction at the highest levels of government public health response.

In the wake of the COVID-19 pandemic, this trend has been followed by calls for infectious disease forecasts to 
be not only designed, but 
also evaluated in ways that align specifically with how forecasts can be used to inform such outbreak control
decisions \citep{marshall2023predictions, bilinski_adaptive_2023}.
This contrasts, however, with the historically standard practice of measuring the quality of disease forecasts using general purpose
accuracy and skill scores, especially those that have implementations available in existing software when the relevant outbreak occurs. For point forecasts,
the root mean square error (RMSE) (e.g., \cite{papastefanopoulos2020covid}) and the mean absolute error (MAE) (e.g., \cite{johansson2016evaluating}) 
are common choices.
For probabilistic forecasts, (which are the focus of this paper), researchers have often relied on the logarithmic score (LS), which
has been used, for example, to evaluate (with respect to both time and space) the skill of US seasonal influenza forecasts \citep{mcgowan_collaborative_2019,reich_collaborative_2019} as well as forecasts targeting surveillance measures of dengue incidence in Peru and Puerto Rico
 \citep{johansson_open_2019}.
More recently, the continuous ranked probability score (CRPS) and a discretized version adapted to multi-quantile forecasts, the weighted interval 
score (WIS) \citep{bracher2021evaluating} (named for its expression in terms of central prediction interval coverage), have gained prominence.
For example, the CRPS was used to assess probabilistic forecasts (based on random effect models) of dengue incidence at the district level 
in Vietnam \citep{colon-gonzalez_probabilistic_2021}. And the WIS has been used during the COVID-19 pandemic (through efforts with which we 
have been intimately involved) to evaluate forecasts of 
observed cases, hospitalizations and deaths in the US and Europe, as reported by municipal, state, and federal surveillance systems 
\citep{cramer_evaluation_2022,fox_real-time_2022,sherratt2023predictive}.

While it should be noted that there are ways to interpret any of these scores abstractly through the lens of decision theory, and that all of 
the application-specific papers cited above benefited from direct collaboration with public health agencies, a key impetus for the
present work has been that we were not able to find in any of them, nor in the literature they represent, explicit connections between how a 
forecast was evaluated and how that forecast was used in practice.

A general phenomenon at play here --- one that has been observed repeatedly over the past few decades in other fields such as finance, supply chain management, and meteorology --- is that while scores such as RMSE, MAE, LS, CRPS, and WIS can describe the \emph{quality} of a forecast 
at corresponding to the observed disease outcome, they will often fail to register the \emph{value} (or lack thereof) 
a forecast has in the context of a specific decision. We refer the reader to \cite{yardley2021utility_cost_forecasts} and the references collected 
therein (especially the foundational \cite{murphy1993whatisagoodforecast}) for a general overview touching on a wide range of forecasting contexts and  
also to \cite{pesaran2002decision_based_eval} for a clear discussion from an econometric perspective.

%------ literature that seems to be aware of the problem we see and how it falls short ------%
This is to say that the quality-value distinction is not a new concept in the larger forecasting community. We are aware, however, of only a 
limited literature attempting to connect the value of infectious disease forecasts to their impact on and through policy.  And
within this body of work we have found the discussion of such a connection to still be at a formally and quantitively imprecise stage. 
In \cite{ioannidis2022forecastingCOVIDfailed}, the possible negative consequences of inaccurate forecasts of infectious disease are discussed, but there is no attempt to quantify the utility or loss incurred as a result of those forecasts. \cite{bilinski_adaptive_2023} explore ways in which predictive classifiers of local COVID-19 risk levels in the US could be tuned to policymaker preferences for different
costs associated with over- and under-reaction to disease dynamics, but they do not clearly identify the source of these costs or how they
depend on quantifiable policy choices. A similar discussion related to dengue countermeasures in Vietnam appears in
\cite{colon-gonzalez_probabilistic_2021}.  A novel version of the WIS informally motivated by utility considerations is devoloped in 
\cite{marshall2023predictions}, but the score is not derived in a decision-theoretic manner.  There is also a thread of literature that frames infectious disease modeling as a component of a larger system for understanding how policy goals, means, and choices interact and constrain
one another. As an example, \cite{Probert2016decisionMakingFootMouth} explore how policy recommendations ought to flow from a possibly 
incongruous set of simulation-based projection models of a hypothetical
foot-and-mouth disease outbreak when there are ranges of plausible responses and stakeholder interests. Decision theory plays a prominent
role here but not explicitly as a way to evaluate the choices made in developing or running the models.

% However, as we show in this paper, it is possible, using standard tools of decision theory, to construct formal methods (detailed in Section 
% \ref{sec:methods.detailed.decisiontheory})) for evaluating probabilistic forecasts that are directly and uniquely tied to policy decisions, 
% in particular those related to resource allocation.  In essence, we will define the score (as a measure of value) of a forecast as  
% Furthermore, we find evidence that such scores can illuminate important features of forecasts related to their value as decision-making inputs which
% are missed by traditional quality (that is, accuracy and skill) measures.

% %------ what you can do instead ------%
% In decision making settings where it is possible to quantify the utility or loss associated with a particular action, standard tools of decision theory provide a procedure for developing forecast scoring rules that measure the value of forecasts through the quality of the decisions that they lead to (see details in Section \ref{sec:methods.detailed.decisiontheory}).

% %In finance, the value of forecasts can often be measured by the profits generated by trading decisions informed by the forecasts, perhaps adjusted for risk levels \cite[e.g.,][]{leitch1991economicForecastEval, cenesizoglu2012returnPredictionEconValue}. In applications to supply chain management and meteorology, the value of forecasts has typically been operationalized by considering the costs associated with decisions regarding the amount of inventory to hold or the level of protection against the impacts of extreme weather events to enact \cite[e.g.,][]{catt2007assessingcostofforecasterror, petropoulos2019inventoryperformanceforecasting, palmer2002economic, pappenberger2015monetarybenefitfloodwarnings}.

% In this framework, a forecast has value if it leads to decisions with low total costs.
% As one example,
% % in supply chain management these decisions may incur costs related to holding inventory, labor, or providing poor service, while
% in meteorology, policy makers need to balance the costs of implementing protective measures with the costs of potentially preventable weather damages \citep{palmer2002economic, pappenberger2015monetarybenefitfloodwarnings}.

%------ our game plan ------%
In this work, we begin to fill this gap between the ways that infectious disease forecasts have traditionally been evaluated and the ways that they have been used to support public health policy. To do so, we consider a setting in which forecasts are used to help determine the allocation of a limited quantity of medical supplies across multiple regions.
In section~\ref{sec:methods} of the paper, we define a new forecast scoring rule --- the {\em allocation score} --- that evaluates forecasts based on how beneficial resource allocations derived from them would turn out to be.
%Unlike the traditional measures of forecast skill such as WIS, CRPS, and log score, the proposed allocation score measures the suitability of a forecast as an input to this specific decision making task.
In section \ref{sec:application}, we present an illustrative analysis using the allocation score to evaluate forecasts of hospital admissions in the US that were made leading up to and during the Omicron wave that peaked in January of 2022.
This analysis is ``synthetic'' insofar as it is not intended to correspond to any specific historical record of allocation decisions 
that could have been supported by hospitalization forecasts during this period.
However, we view the general allocation problem on which our framework is based as a versatile template for 
formalizing real-world decisions that must constantly be made in real-time by public health administrators around the globe, especially those
related to hospital capacity, ventilator usage, doses needed for specific medications and other situations where an outbreak creates sudden and
highly variable demand for potentially scarce resources.
